{{< include header.qmd >}}

---
execute:
  echo: true
---

## Future sessions {.smaller}

```{r}
#| echo: false

KINDR::training_sessions(tr_type = "R", start_date = params$date, n = 5)
```

## Session outline

This session is an 🌶🌶 **intermediate practical** designed for those with some R experience. The aim of this session is to do three things with dplyr:

+ explain the difference between binds and joins
+ to demonstrate a range of different joining strategies
+ to explain dplyr's new tools for controlling joins (`.by`, `by()`, `multiple()`, `unmatched()`)

. . .

You might also like some of the other dplyr-themed practical sessions:

* [Exploring data with dplyr](exploring_data_with_dplyr)
* [Summarising data with dplyr](dplyr_group_summarise.qmd)
* [Tidyselect](tidyselect.qmd)

{{< include "content/Joining data with dplyr.qmd" >}}

## Future sessions {.smaller}

```{r}
#| echo: false
KINDR::training_sessions(tr_type = "R", start_date = params$date, n = 5)
```

{{< include footer.qmd >}}