{{< include header.qmd >}}

## Welcome

+ this session is for 🌶🌶 intermediate users
+ if you can't access the chat, you might need to join our Teams channel: **tinyurl.com/kindnetwork**
+ this session can be followed practically - you'll need Power BI desktop ([plus the sample file](https://github.com/NES-DEW/KIND-training/raw/main/bi_training/data/ae_weeks_start.pbix)) to do that

## The KIND network

-   a social learning space for staff working with **k**nowledge, **i**nformation, and **d**ata across health, social care, and housing in Scotland
-   we offer social support, free training, mentoring, community events, ...
-   [Teams channel](https://teams.microsoft.com/l/team/19%3AQZ7-PbFVcziG2piHLt1_ifey3I2cwFL0yBuTSS8vVao1%40thread.tacv2/conversations?groupId=106d08f3-9026-40e2-b3c7-87cd87304d58&tenantId=10efe0bd-a030-4bca-809c-b5e6745e499a) / [mailing list](https://forms.office.com/pages/responsepage.aspx?id=veDvEDCgykuAnLXmdF5JmpopIZB9ynRJnrPUHVFccipURjM2NkZJUkhGOFlQRjQxRFhVUTgwT0UwVyQlQCN0PWcu)

## Future sessions {.smaller}

```{r}
KINDR::training_sessions(tr_type = "Power BI", start_date = params$date, n = 5)

```


## Session outline

+ DAX
+ measures
  + Excel measures
  + measures vs calculated columns
+ basic measures for data summary
+ more complex measures for re-filtering

## DAX

+ **d**ata **a**nalysis e**x**pressions
+ based on Excel formula language
+ been in use for about 15 years, initially in Excel

## Let's try one in Excel

+ [sample Excel](https://github.com/NES-DEW/KIND-training/raw/main/bi_training/data/excel_measures.xlsx)
+ PivotTable plus data model needed
+ Measures live in tables
+ written in DAX: `=CONCATENATEX(VALUES(Table134[Fluid spilled on patients]), Table134[Fluid spilled on patients], ",")`

## Into Power BI

+ [sample file](https://github.com/NES-DEW/KIND-training/raw/main/bi_training/data/ae_weeks_start.pbix)
+ confusingly, we find DAX in two different places...

## Calculated columns

+ Let's calculate an average
+ `aver_att = AVERAGE(ae_activity[att])`

## Measures

+ dynamically recalculate depending on filter context
  + so need to live on a card
+ so same DAX = different result
+ `aver_att_meas = AVERAGE(ae_activity[att])`

. . .

## Basic functions for measures

+ `SUM`, `AVERAGE`, `COUNT`, `MAX`
+ measure-in-measure
+ `&` and `ROUND`

## Row-wise operations
+ `SUMX`
+ `CONCATENATEX` (from Excel example)

## More interesting measures: filter

+ `count_12 = COUNTROWS(FILTER(ae_activity, ae_activity[over12] > 20))`
+ `FILTER` allows us to look at how parts of our data change

## More interesting measures: calculate

+ `count_12_calc = CALCULATE(COUNT([att]), ae_activity[over12] > 20)`
+ `CALCULATE` changes the default filter context

## More interesting measures: keepfilters

+ `count_12_calc_kf = CALCULATE(COUNT([att]), KEEPFILTERS(ae_activity[over12] > 20))`
+ `KEEPFILTERS` pokes the default filter context through into a `CALCULATE`

## Resources

+ [Definitive Guide to DAX](https://www.waterstones.com/book/definitive-guide-to-dax-the/marco-russo/alberto-ferrari/9781509306978)
+ [SQLBI CALCULATE](https://www.youtube.com/watch?v=HxZLkmpY6BA)
+ [SQLBI KEEPFILTERS](https://www.youtube.com/watch?v=Cj6rI0fRCaQ)

## Future sessions {.smaller}

```{r}
KINDR::training_sessions(tr_type = "Power BI", start_date = params$date, n = 5)

```

{{< include footer.qmd >}}
