{{< include header.qmd >}}

## Future sessions {.smaller}

```{r}
KINDR::training_sessions(tr_type = "Power Platform", start_date = params$date, n = 5)
```


## Welcome

- this session is a non-technical overview
- we'll introduce Power Query, run through some examples, and have plenty of time for chat
- as ever, this is social learning. Cameras on, mics off, if possible please!



## Session outline

- data wrangling
-   what's Power Query, and why this session?
-   strengths and weaknesses
-   skill development

## Data wrangling

- data wrangling = the work required to get data in the right shape for analysis
- often manual and repetitious 
- e.g. some staff spend 90% of their time doing the same Excel work
- this is **almost certainly** the result of poor practices, rather than the tech
- (unsolicited advice): we are far too accepting of that drudgery

## The golden rule {auto-animate=true}

::: {style="margin-top: 100px;"}
Don't repeat yourself
:::

## The golden rule {auto-animate=true}

::: {style="margin-top: 200px; font-size: 2.5em; color: gold;"}
Don't repeat yourself
:::

## Power Query

- specialist data wrangling toolkit
- allows you to automate data wrangling
- available in Excel from c.2013
    -   add-ins from Excel 2010/3
    -   built-in from Excel 2016
-   also found in Power BI (and some of the Microsoft SQL products)

## Why this session?

- (opinion) PQ is criminally under-used
- if you're repeating yourself in Excel, consider doing it in PQ
- benefits beyond automation

## Workflow

-   load data into PQ from many sources:
  - tables, spreadsheets, CSV, SQL, ...
-   process that data by building a **transformation pipeline**
-   return your data to Excel
- potentially, run that same pipeline again when the source data changes

## Sample data

+ some Excel tables
+ [GP practice data](https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/f3633113-9768-4676-8ed1-6695fe385bbe/download/practice_contactdetails_apr2025.csv)
+ [Open Annual inpatient/daycase data](https://www.opendata.nhs.scot/dataset/9e032286-e5bf-4eb0-837d-c0a05fe0d839/resource/32058e0a-0107-4035-891c-5c6a8571415a/download/inpatient_and_day_case_by_council_area_age_sex.csv)

## Strengths

- flexible: do almost anything to your data
- reproducible: do the same operations over many files
- powerful: has advanced tools for reshaping data

## Weaknesses

- harder to learn than Excel
- requires some thought and set-up
- poor documentation
- bafflingly small user-base

## Future sessions {.smaller}

```{r}
KINDR::training_sessions(tr_type = "Power Platform", start_date = params$date, n = 5)
```

{{< include footer.qmd >}}